Evaluating machine translation for assimilation via a gap-filling task

### 2015-03-06 2015-eamt-assim/2015-eamt-assim.tex
Final draft
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@59235 72bbbca6-d526-0410-a7d9-f06f51895060

\maketitle
\renewcommand{\baselinestretch}{0.97} % the undetectable 3% squeeze
\comment{EA: has not it first been used in trosterud12?}
\begin{abstract}
  This paper provides additional observations on the viability of a
  strategy independently proposed in 2012 and 2013 for evaluation of
  machine translation for assimilation purposes. The strategy involves
  human evaluators, who are asked to restore key words (to fill gaps) in
  reference translations. The evaluation method is applied to two translation
  directions.\comment{MLF: directions \(\to\) language pairs?} To reduce the amount of time required to prepare tasks
  and analyze results, an open-source task management system is
  introduced. The evaluation results suggest that the gap-filling task
  reflects users' understanding of text,\comment{MLF: ``suggests''? I think this is what we assume to interpret our results: how about ``shows''? We don't show correlation with other evaluation.} and may be used to measure MT
  quality for assimilation purposes.
\end{abstract}

@@

\comment{EA: I'd like a citation for the ``most massive application'', could you suggest something? It is from Mikel's paper, but no reference there -- should I cite Mikel?}
\comment{MLF: I added a footnote, please reduce and move citation to the reference}
Despite the fact that, as a result of widespread usage of online MT, assimilation (or gisting) is currently the most frequent application of MT,\footnote{“In a given day we [Google] translate roughly as much text as you’d find in 1 million books. To put it another way: what all the professional human translators in the world produce in a year, our system translates in roughly a single day.”, Franz J. Och, \url{http://tinyurl.com/och-gisting}} few methodologies are established for assimilation evaluation of MT. The methods include post-editing and comparison by bilingual experts \citep{ginesti09},\comment{MLF: should we also cite the original WMT reference cited by Ginestí et al.?} and multiple choice tests \citep{jones07,trosterud12}. These approaches are often costly and prone to subjectivity: see the discussion in \cite{oregan13}. As an alternative, the modification of \emph{cloze}\comment{\emph{cloze} stands for \emph{closure}.} testing \citep{taylor53} was introduced for assimilation evaluation, first by \citep{trosterud12} as a supplementary, and then by \citep{oregan13} as a stand-alone method. Prior to this, cloze tests have been used to evaluate raw MT quality \citep{vanslype79,somers00}. While these authors ask informant to fill gaps in MT output, \citet{trosterud12} and \citet{oregan13} propose to fill gaps in the reference (human) translation. The gap-filling task models how well users comprehend the key points of the text, as it is roughly equivalent with answering questions.\comment{MLF: check if you agree with this} Thus, the method does not directly evaluate the quality of machine-produced text, but rather its usefulness in understanding the meaning of the original text.

\comment{EA: I understand we selected tat-rus and eng-kaz because they were in development. why did we select eu-es?}
The gap-filling method has been successfully used to evaluate the Basque--English Apertium language pair. In this work we extend the evaluation to two more language pairs: Basque--Spanish and Tatar-Russian. For the latter pair, the evaluation served as a quality check in the period of active development during the Google Summer of Code 2014 programme. In addition to evaluating, we explore new aspects of the experiment: the correlation between evaluators' scores, and the influence of the texts' linguistic domain. To facilitate the evaluation, we introduce an automated system which creates task sets from parallel corpora given a range of parameters (number of gaps in a sentence, hint type, gap filler,\comment{MLF: gap filler or gap marker?} etc.), checks evaluators' answers, and calculates and reports generalized results. This system is integrated into the Appraise MT evaluation platform \citep{federmann12}; the code is open source and is available on GitHub.\footnote{\phantom{\url{https://github.com/Sereni/Appraise}}}\comment{MLF: check if phantom URL is ok}

We anticipate that the assessed MT systems will contribute to the users' understanding of text, that is, the evaluation will show better results in MT-assisted task type.\comment{MLF: this last sentence is very hard to follow} We also expect to see different results depending on such parameters as text domain and the number of gaps in a sentence in relation to sentence length.

@@

\begin{table*}
  \begin{tabular}{|l|p{15cm}|}
%@@@  \textbf{Src (Tatar)} & \rus{???? ?????? ???????; ????? 7 ??????????? ??? ??????? ?????, ????? ??? ????? 100 ???? ??????? ???????.}
%  \hline
%  \textbf{MT (Russian)} & \rus{Половина сказать часы; вам 7 тукталыштан поздний опуститься нужный , после его снова 100 метр сказать идти.} \\
     \hline
     \textbf{Ref.} & \rus{Примерно полчаса; вам нужно выйти через 7 остановок, потом пройти ещё около 100 метров.} \\\hline
     \textbf{10\%}   & \rus{Примерно полчаса; вам нужно выйти через 7 \{ \}, потом пройти ещё около 100 метров.} \\
     \textbf{20\%}   & \rus{\{ \} полчаса; вам нужно \{ \} через 7 остановок, \{ \} пройти ещё около 100 метров.} \\
     \textbf{30\%} & \rus{Примерно полчаса; вам нужно \{ \} через 7 \{ \}, потом пройти \{ \} около 100 \{ \}.} \\
     \hline
  \end{tabular}
  \caption{Example of different gap percentage settings for a Russian reference sentence.} 
  \label{table:percentage}
\end{table*}
\comment{MLF the gap percentage table 2 is not referenced in the text; also, i have changed it to have the \textbf{Ref} row in the table?}

This section discusses the reasoning behind the gap-filling method and task structure. The gap-filling method 
of evaluating machine translation for assimilation purposes is based on the following hypothesis: a reader's understanding 
of a given text correlates with the number of words they are able to correctly restore in the text. Therefore, the 
base of an assimilation task is a (reference) sentence, where some of the words are blacked out, or removed. The sentence \comment{EA: I'm sorry, I'm not sure what else I could explain here...}
is produced by a human (as opposed to machine-translated), and it is in the language known to evaluators, which is also 
the \emph{target language} of the machine translation system.
The additional elements of the task are what we call hints, or extra sentences that help the participant to understand 
the main sentence. There are two types of hints: first, the \emph{source}, which is semantically equivalent to the \emph{reference}, 
also human-produced, but in the source language of the pair. The second type is the \emph{machine-translated} hint, which 
comes from the machine translation of the \emph{source} sentence. Table~\ref{table:example} shows a sample task, and Figure~\ref{figure:screenshot} shows the task in the online evaluation environment.\comment{Use the highest possible resolution for the figure so that aliasing is imperceptible}

\begin{figure*}
  \centering
\includegraphics[width=0.9\textwidth,resolution=144]{appraise-scr}
 \caption{An example set of sentences in an online environment. The task is Russian legal text with 30\% gaps.}
\label{figure:screenshot}
\end{figure*}

\comment{EA: I don't really have a good explanation why we used source+mt tasks, except for being exhaustive. could you suggest anything?}
In the course of the experiment, following \cite{oregan13}, the following hint combinations are offered:
\begin{description}
\item[Reference sentence only:] The participants are asked to fill the gaps without being given
any context. This task serves as a baseline score and as an indicator of gaps
that can be completed using common knowledge or language intuition (e.g.
idioms and strong collocations).
\item[Reference sentence and source sentence:] By setup, the
participants have no command of the source language, however, it may help them to fill
in proper nouns or loan words.
\item[Reference sentence and MT hint:] In addition to the task, the
participants see the source sentence translated via the MT system Apertium \citep{forcada11}. This type of task is
used for measuring the contribution of machine translation to understanding the
gist of text.
\item[Reference sentence and both hints:] This task is
added for completing the possible combination of task types, and to check
for unexpected insights, should they arise.
\end{description}

In order to prepare the evaluation questions, we determine and remove the keywords\comment{MLF: decide, \emph{key words} or \emph{keywords} everywhere} in the 
reference sentences. We consider two parameters: the list of allowed parts of speech (PoS), and the 
number of gaps relative to sentence length (``gap density''). For the evaluations described in 
this paper we use gap densities of 10, 20 and 30 percent, and the following parts of speech: noun, 
proper noun, adjective, adverb and lexical verb (as opposed to auxiliary verb).

For each sentence, the list of candidate keywords is prepared. It is composed of all the words that 
fall into the allowed PoS list. The number of gaps in the sentence is calculated based on 
sentence length and specified gap density. All reference sentences are longer than 10 words. Finally, the required number of keywords is selected 
from the candidate list in such a manner that the gaps are distributed evenly throughout the sentence. We start at a random word in a sentence and check whether it is a keyword. If yes, we remove it, and move \emph{n} words forward, going back to the beginning of sentence if necessary. The step length \emph{n} is the sentence length divided by the desired number of gaps. If the word is not a keyword, or has already been removed, we look at the next word instead. The process is repeated until the designated number of words has been removed, or until there are no more words in the keyword list.\comment{MLF: O'Regan and Forcada did something lots simpler: why not here? It is probably a bit late to change this}

Keyword removal is one of the most time-consuming steps in task preparation. It normally requires human effort, because we would like to determine the words that contribute the most to understanding the text as opposed to removing random words. \comment{MLF: the procedure described so far seems automatic. What's manual and time consuming?} In our setup, the above procedure is performed by a script integrated into the task generation pipeline. Parts of speech are determined with Apertium morphological analyser. To control for homonymy, we only allow the word into the candidate list if all of its possible part of speech attributions are on the PoS list. For example, if we only allow nouns on the word list, and the word "fly" receives two possible part of speech attributions from the tagger, noun and verb, it is not considered for the candidate list.

@@

The participants are given the instructions in their native language; these instructions are repeated above each task in the evaluation system. For the participants' convenience, the body of questions is split into smaller groups which allow indefinite time breaks\comment{MLF: time break OK? Possibility: \emph{multiple sessions}.} between them. The instructions are the 
following: read all the available hints and fill each gap with one suitable word, guessing if unsure.
Participants' answers are recorded and marked correct or incorrect automatically. In
addition, the time taken to fill the gaps in one sentence is recorded.

This variety of the gap-filling task requires open answers, and it is therefore possible that the participants may
provide words that fit the gaps well, but do not match the original answer. To account for
these cases, we process all the answers to detect possible synonyms (a method suggested by \cite{oregan13}). An answer is
considered a candidate synonym if it is given by two or more evaluators, and it does not match
the answer word. We record each candidate synonym along with the answer key and the
context sentence (see Table~\ref{table:syn} for an example). Based on this data, a native speaker of the target language decides
whether the candidate synonym is an acceptable replacement to the answer key in the given
context. We then check participants' results against the compiled synonym list and
increase scores where appropriate. On average, the scores improved by three percent in all evaluation modes. Candidate synonyms are extracted automatically from the evaluators' responses,  and each individual score is automatically updated according to the synonym list.

\begin{table*}
  \centering
  \begin{tabular}{|l|l|}
     \hline
     \textbf{Sentence:}   & Aprender a jugar y divertirse en el agua sin asumir riesgos. \\
     \textbf{Key:}   & asumir \\
     \textbf{Synonym:} & correr \\
     \hline
  \end{tabular}
  \caption{An example candidate synonym. The original word \emph{asumir} may be replaced with \emph{correr} in this context.} 
  \label{table:syn}
\end{table*}
\comment{EA: added}
The synonym lists for Basque--Spanish and Tatar--Russian contain 52 and 25 words, respectively. Time taken to compile each list depends on the number of candidate synonyms, and in our case was approximately 30 minutes. 
\comment{FMT: How long did making the synonym list take, how many synonyms were added? More details here. Also perhaps an example?}

@@

% a few notes on time
The latter trend holds: the average time taken to fill the gaps increases and the average percentage of correct answers decreases as the relative number of gaps goes up. The larger number of gaps in the sentence makes it more
difficult to predict the answer based on the context, and also leaves more room for
mistakes. Exploring different percentage-mode combinations, we may note that the 10\% no-hint tasks take the least time to complete. We would have expected longer completion time, since the participant must come up with their own answer unassisted. However, in the no-hint task the participant is required to read only one (reference) sentence, as opposed to two or three (reference and hints) in other tasks. Also, the number of gaps in 10\%-gap tasks is low, as it never exceeds three. We found that, as opposed to trying to devise the best word for no-hint gaps, the participants often resorted to filing these gaps with random words, which takes little time.
\comment{EA: anything else I could write about times?}

@@

Two records break this trend: the no-hint 10\% sentences in Basque--Spanish, which scored significantly higher than the source-hint in the same category, and MT+source 10\% sentences in Tatar--Russian, which we would have expected to score higher than the corresponding MT task. In the first case, this is largely due to the use of synonyms list. Before taking synonyms into account, the scores were 32 and 35 percent for source and no-hint tasks, respectively. This still shows a small difference in favor of no-hint tasks. However, the latter percentage increases significantly after we extend the answer list with synonyms. Such an increase suggests that, in this case, the content words were restored by 
\comment{EA: to JA: I am not sure if this qualifies as common sense. What I meant is that people correctly guessed the meaning of the word in place, and expressed it in their own manner. Changed it to semantic context, is this a better rephrasing?}
semantic context rather than through strong collocation. The second pattern, low scores in Tatar--Russian 10\% MT+source, does not stem from the task content. Instead, it is the result of the fixed order of tasks: the participants have always been given MT+source 10\% sentences first, followed by other task types.\comment{MLF: Why? Not random? \cite{oregan13} did random! Why was this changed?} The participants have not received any training tasks before the main evaluations. Therefore, it is possible that the accommodation period is responsible for lower-than-expected scores in this mode of evaluation.

@@

\begin{table*}
\centering
  \begin{tabular}{|l|r|r|r|r|}
  \hline
 \multirow{2}{*}{\textbf{Density}} & \multicolumn{4}{|c|}{Evaluation mode} \\\cline{2-5}
                                            & \textbf{MT \& Src} & \textbf{MT} & \textbf{Src} & \textbf{No hint} \\
\hline
10\%&0.598&0.459 &0.711 &0.517 \\
20\%&0.740&0.667 &0.473 &0.261 \\
30\%&0.534&0.581 &0.411 &0.412 \\
\hline
\end{tabular}
\caption {Tatar--Russian: Krippendorff Alpha measure} 
\label{table:alpha-tat-rus}
\end{table*}
We selected this measure because of its compatibility with more than two annotators per task and the missing data (not all the gaps were evaluated). To calculate Krippendorff's alpha we used an algorithm implementation by Thomas Grill,\footnote{http://grrrr.org}\comment{MLF: Give a direct reference: I could not find the algorithm in such a wide website. Use tinyurl if it is too long.} dividing the answers in each gap into two categories: correct and incorrect. The previously obtained synonym lists were taken into account, i.e. if the two answers are different but both correct, they fall into one category. The measure was calculated separately for each hint and percentage combination.

As the MT and MT+source hints are introduced, the agreement between annotators increases (measures closer to 1): the annotators are more consistently correct or incorrect in each given sentence.\comment{EA: I cite Juan Antonio here, because I am not completely sure what alpha shows about our experiment (lack of knowledge on my part)}  The agreement measure for the same sentences without hints is closer to zero, which contributes to the reliability of our methodology. We note the outlier score in Tatar--Russian 10\% source tasks, which has the most contribution from the news texts (agreement of 0.917). This set of sentences contains many loan words, which have similar form in Tatar and Russian (e.g. president, minister, championship), and are understood by Russian speakers. The gaps with loan words have mostly been filled correctly, while there was some disagreement in other gaps.

% a little on categories
\subsection{Results for different domains}
\comment{EA: and what would we like to say about those? The results depend heavily on the Tatar--Russian system. Do we have to provide info on what vocabulary it handles better, what it has and has not been trained on? Or do we just say that casual texts are easier to fill than legal texts?}
For the Tatar--Russian language pair the participants were offered texts from three different domains (in equal proportions): casual conversations, legal texts and news. The results by domains are displayed in tables \ref{table:rus-casual}, \ref{table:rus-legal}, and \ref{table:rus-news}. The MT system used in the evaluation has been targeted to translate texts from all three of the domains. Taking into consideration the above discussion of 10\% MT+Source tasks, we observe similar results across the three categories. Note that, in casual and news texts, the source sentences paired with MT significantly improve participants' performance compared to MT-only task mode. This may be due to the fact that many words are borrowed from Russian into Tatar, and are in fact understood by Russian speakers.

@@

\begin{table*}
\centering
  \begin{tabular}{|l|r|r|r|r|}
  \hline
 \multirow{2}{*}{\textbf{Density}} & \multicolumn{4}{|c|}{Evaluation mode} \\\cline{2-5}
                                            & \textbf{MT \& Src} & \textbf{MT} & \textbf{Src} & \textbf{No hint} \\
\hline
10\%&62.19 \(\pm\) 32.07&58.04 \(\pm\) 27.96&39.90 \(\pm\) 38.96&48.79 \(\pm\) 39.84\\
20\%&64.81 \(\pm\) 30.03&70.18 \(\pm\) 26.98&31.37 \(\pm\) 28.42&30.91 \(\pm\) 30.22\\
30\%&48.13 \(\pm\) 26.08&39.55 \(\pm\) 23.90&25.94 \(\pm\) 19.67&18.07 \(\pm\) 18.15\\
\hline
\end{tabular}
\caption {Basque--Spanish: Average number of gaps successfully filled (\%), using a synonym list} \label{table:res-eus-spa}
\end{table*}

\begin{table*}
\centering
  \begin{tabular}{|l|r|r|r|r|}
    \hline
    \multirow{2}{*}{\textbf{Density}} & \multicolumn{4}{|c|}{Evaluation mode} \\\cline{2-5}
                                            & \textbf{MT \& Src} & \textbf{MT} & \textbf{Src} & \textbf{No hint} \\
\hline
10\%&63.76 \(\pm\) 41.15&57.02 \(\pm\) 41.57&53.52 \(\pm\) 43.57&45.96 \(\pm\) 40.86\\
20\%&59.77 \(\pm\) 33.10&64.93 \(\pm\) 30.95&45.60 \(\pm\) 31.23&38.56 \(\pm\) 32.11\\
30\%&55.83 \(\pm\) 26.13&58.77 \(\pm\) 28.07&40.00 \(\pm\) 28.16&35.45 \(\pm\) 29.76\\
    \hline
  \end{tabular}
  \caption{Tatar--Russian: Average number of gaps successfully filled (\%), using a synonym list}
  \label{table:res-tat-rus}
\end{table*}

@@

\centering
\begin{tabular}{|r |*{4}{c}|}
\hline
  &\multicolumn{4}{c|}{\textbf{Evaluation mode}}\\
\hline
\textbf{Gap percentage} & \textbf{MT \& Src} & \textbf{MT} & \textbf{Src} & \textbf{No hint} \\
10\%&63.64 \(\pm\) 44.84&63.64 \(\pm\) 43.12&62.12 \(\pm\) 46.11&53.03 \(\pm\) 44.28\\
20\%&73.33 \(\pm\) 31.51&63.33 \(\pm\) 35.85&41.11 \(\pm\) 27.88&37.78 \(\pm\) 30.94\\
30\%&70.28 \(\pm\) 31.12&59.58 \(\pm\) 23.89&39.44 \(\pm\) 27.10&37.71 \(\pm\) 33.32\\
\hline
\end{tabular}
\caption {Tatar--Russian, casual conversations: Average number of gaps successfully filled (\%), using a synonym list} \label{table:rus-casual} 
\end{table*}

>>>>

\maketitle
\renewcommand{\baselinestretch}{0.97} % the undetectable 3% squeeze
\begin{abstract}
  This paper provides additional observations on the viability of a
  strategy independently proposed in 2012 and 2013 for evaluation of
 MT for assimilation purposes. The strategy involves
  human evaluators, who are asked to restore keywords (to fill gaps) in
  reference translations. The evaluation method is applied to two language pairs, Basque--Spanish and Tatar--Russian. To reduce the amount of time required to prepare tasks
  and analyse results, an open-source task management system is
  introduced. The evaluation results show that the gap-filling task
  reflects users' understanding of text, and may be used to measure MT
  quality for assimilation purposes.
\end{abstract}

@@

Despite the fact that, as a result of widespread usage of online MT, assimilation (or gisting) is currently the most frequent application of MT (in 2012, daily output of Google Translate matched the yearly output of human translations \citep{google}), few methodologies are established for assimilation evaluation of MT. The methods include post-editing and comparison by bilingual experts \citep{ginesti09},\comment{MLF: should we also cite the original WMT reference cited by Ginestí et al.?} and multiple choice tests \citep{jones07,trosterud12}. These approaches are often costly and prone to subjectivity: see the discussion in \cite{oregan13}. As an alternative, the modification of \emph{cloze}\comment{\emph{cloze} stands for \emph{closure}.} testing \citep{taylor53} was introduced for assimilation evaluation, first by \citet{trosterud12} as a supplementary, and then by \citet{oregan13} as a stand-alone method. Prior to this, cloze tests have been used to evaluate raw MT quality \citep{vanslype79,somers00}. While these authors ask informants to fill gaps in MT output, \citet{trosterud12} and \citet{oregan13} propose to fill gaps in the reference (human) translation. The gap-filling task models how well users comprehend the key points of the text, as it is roughly equivalent with answering questions.\comment{MLF: check if you agree with this} Thus, the method does not directly evaluate the quality of machine-produced text, but rather its usefulness in understanding the meaning of the original text.

The gap-filling method has been successfully used to evaluate the Basque--English Apertium language pair. In this work we extend the evaluation to two more language pairs: Basque--Spanish and Tatar-Russian. The former pair, while not producing the output suitable for post-editing, is a good example of an assimilation MT system. In addition, Basque and Spanish are not mutually understandable, and therefore constitute a good pair for evaluation. For the latter pair, the evaluation served as a quality check in the period of active development during the Google Summer of Code 2014 programme. In addition to evaluating, we explore new aspects of the experiment: the correlation between evaluators' scores, and the influence of the texts' linguistic domain. To facilitate the evaluation, we introduce an automated system which creates task sets from parallel corpora given a range of parameters (number of gaps in a sentence, hint type, gap filler,\comment{MLF: gap filler or gap marker?} \comment{EA: well, initially we had questions with lemmas in place of gaps, and also multiple choice questions where answer options got generated automatically. This is not used in this evaluation, but it's still in the toolkit. Should I remove?} etc.), checks evaluators' answers, and calculates and reports generalized results. This system is integrated into the Appraise MT evaluation platform \citep{federmann12}; the code is open source and is available on GitHub.\footnote{\phantom{\url{https://github.com/Sereni/Appraise}}}\comment{MLF: check if phantom URL is ok}

We anticipate that the assessed MT systems will contribute to the users' understanding of text, that is, the users will show better results in tasks when assisted with MT. We also expect to see different results depending on text domain and the relative number of gaps in a sentence.

@@

\begin{table*}
  \begin{tabular}{|l|p{15cm}|}
%@@@  \textbf{Src (Tatar)} & \rus{???? ?????? ???????; ????? 7 ??????????? ??? ??????? ?????, ????? ??? ????? 100 ???? ??????? ???????.}
%  \hline
%  \textbf{MT (Russian)} & \rus{Половина сказать часы; вам 7 тукталыштан поздний опуститься нужный , после его снова 100 метр сказать идти.} \\
     \hline
     \textbf{Ref} & \rus{Примерно полчаса; вам нужно выйти через 7 остановок, потом пройти ещё около 100 метров.} \\\hline
     \textbf{10\%}   & \rus{Примерно полчаса; вам нужно выйти через 7 \{ \}, потом пройти ещё около 100 метров.} \\
     \textbf{20\%}   & \rus{\{ \} полчаса; вам нужно \{ \} через 7 остановок, \{ \} пройти ещё около 100 метров.} \\
     \textbf{30\%} & \rus{Примерно полчаса; вам нужно \{ \} через 7 \{ \}, потом пройти \{ \} около 100 \{ \}.} \\
     \hline
  \end{tabular}
  \caption{Example of different gap percentage settings for a Russian reference sentence.} 
  \label{table:percentage}
\end{table*}
\comment{MLF the gap percentage table 2 is not referenced in the text; also, i have changed it to have the \textbf{Ref} row in the table?}

This section discusses the reasoning behind the gap-filling method and task structure. The gap-filling method 
of evaluating machine translation for assimilation purposes is based on the following hypothesis: a reader's understanding 
of a given text correlates with the number of words they are able to correctly restore in the text. Therefore, the 
base of an assimilation task is a (reference) sentence, where some of the words are blacked out, or removed. The sentence 
is produced by a human (as opposed to machine-translated), and it is in the language known to evaluators, which is also 
the \emph{target language} of the machine translation system.
The additional elements of the task are what we call hints, or extra sentences that help the participant to understand 
the main sentence. There are two types of hints: first, the \emph{source}, which is semantically equivalent to the \emph{reference}, 
also human-produced, but in the source language of the pair. The second type is the \emph{machine-translated} hint, which 
comes from the machine translation of the \emph{source} sentence. Table~\ref{table:example} shows a sample task, and Figure~\ref{figure:screenshot} shows the task in the online evaluation environment.\comment{Use the highest possible resolution for the figure so that aliasing is imperceptible}

\begin{figure*}
  \centering
\includegraphics[width=0.9\textwidth,resolution=144]{appraise-scr}
 \caption{An example set of sentences in the online environment. The task is Russian legal text with 30\% gaps.}
\label{figure:screenshot}
\end{figure*}

In the course of the experiment, following \cite{oregan13}, we offer these hint combinations:
\begin{description}
\item[Reference sentence only:] The participants are asked to fill the gaps without being given
any context. This task serves as a baseline score and as an indicator of gaps
that can be completed using common knowledge or language intuition (e.g.
idioms and strong collocations).
\item[Reference sentence and source sentence:] By setup, the
participants have no command of the source language, however, it may help them to fill
in proper nouns or loan words.
\item[Reference sentence and MT hint:] In addition to the reference sentence, the
participants see the source sentence translated via the MT system Apertium \citep{forcada11}. This type of task is
used for measuring the contribution of machine translation to understanding the
gist of text.
\item[Reference sentence and both hints:] This task is
added to check whether MT and source provide complementary hints.
\end{description}

In order to prepare the evaluation questions, we determine and remove the keywords in the 
reference sentences. We consider two parameters: the list of allowed parts of speech (PoS), and the 
number of gaps relative to sentence length (``gap density''). For the evaluations described in 
this paper we use gap densities of 10, 20 and 30 percent (Table~\ref{table:percentage}), and the following parts of speech: noun, 
proper noun, adjective, adverb and lexical verb (as opposed to auxiliary verb).

For each sentence, the list of candidate keywords is prepared. It is composed of all the words that 
fall into the allowed PoS list. The number of gaps in the sentence is calculated based on 
sentence length and specified gap density. All reference sentences are longer than 10 words. Finally, the required number of keywords is selected 
from the candidate list in such a manner that the gaps are distributed evenly throughout the sentence. We start at a random word in a sentence and check whether it is a keyword candidate. If yes, we remove it, and move \emph{n} words forward, going back to the beginning of sentence if necessary. The step length \emph{n} is the sentence length divided by the desired number of gaps. If the word is not a keyword, or has already been removed, we look at the next word instead. The process is repeated until the designated number of words has been removed, or until there are no more words in the keyword list.\comment{MLF: O'Regan and Forcada did something lots simpler: why not here? It is probably a bit late to change this}

Keyword removal could be one of the most time-consuming steps in task preparation. It normally requires human effort, because we would like to determine the words that contribute the most to understanding the text as opposed to removing random words. \comment{MLF: the procedure described so far seems automatic. What's manual and time consuming?} In our automatic setup, the above procedure is performed by a script integrated into the task generation pipeline. Parts of speech are determined with Apertium morphological analyser. To control for homonymy, we only allow the word into the candidate list if all of its possible part of speech attributions are on the PoS list. For example, if we only allow nouns on the word list, and the word "fly" receives two possible part of speech attributions from the tagger, noun and verb, it is not considered for the candidate list.

@@

The participants are given the instructions in their native language; these instructions are repeated above each task in the evaluation system. For the participants' convenience, the body of questions is split into smaller groups which allow multiple evaluation sessions. The instructions are the 
following: read all the available hints and fill each gap with one suitable word, guessing if unsure.
Participants' answers are recorded and marked correct or incorrect automatically. In
addition, the time taken to fill the gaps in one sentence is recorded.

This variety of the gap-filling task requires open answers, and it is therefore possible that the participants may
provide words that fit the gaps well, but do not match the original answer. To account for
these cases, we process all the answers to detect possible synonyms (a method suggested by \cite{oregan13}). An answer is
considered a candidate synonym if it is given by two or more evaluators, and it does not match
the answer word. We record each candidate synonym along with the answer key and the
context sentence. For example, the word \emph{asumir} is the original answer in the Spanish sentence \emph{Aprender a jugar y divertirse en el agua sin asumir riesgos}. However, two or more evaluators gave a different answer, \emph{corder}. Based on this data, a native speaker of the target language decides
whether the candidate synonym is an acceptable replacement to the answer key in the given
context. We then check participants' results against the compiled synonym list and
increase scores where appropriate. On average, the scores improved by three percent in all evaluation modes. Candidate synonyms are extracted automatically from the evaluators' responses,  and each individual score is automatically updated according to the synonym list.

The synonym lists for Basque--Spanish and Tatar--Russian contain 52 and 25 words, respectively. Time taken to compile each list depends on the number of candidate synonyms, and in our case was approximately 30 minutes.

@@

% a few notes on time
The latter trend holds: the average time taken to fill the gaps increases and the average percentage of correct answers decreases as the relative number of gaps goes up. The larger number of gaps in the sentence makes it more
difficult to predict the answer based on the context, and also leaves more room for
mistakes. Exploring different percentage-mode combinations, we may note that the 10\% no-hint tasks take the least time to complete. We would have expected longer completion time, since the participant must come up with their own answer unassisted. However, in the no-hint task the participant is required to read only one (reference) sentence, as opposed to two or three (reference and hints) in other tasks. Also, the number of gaps in 10\%-gap tasks is low, as it never exceeds three. We found that, as opposed to trying to devise the best word for no-hint gaps, the participants often resorted to filing these gaps with random words, which takes little time.

@@

Two records break this trend: the no-hint 10\% sentences in Basque--Spanish, which scored significantly higher than the source-hint in the same category, and MT+source 10\% sentences in Tatar--Russian, which we would have expected to score higher than the corresponding MT task. In the first case, this is largely due to the use of synonyms list. Before taking synonyms into account, the scores were 32 and 35 percent for source and no-hint tasks, respectively. This still shows a small difference in favor of no-hint tasks. However, the latter percentage increases significantly after we extend the answer list with synonyms. Such an increase suggests that, in this case, the content words were restored by 
semantic context rather than through strong collocation. The second pattern, low scores in Tatar--Russian 10\% MT+source, does not stem from the task content. Instead, it is the result of the fixed order of tasks: the participants have always been given MT+source 10\% sentences first, followed by other task types.\comment{MLF: Why? Not random? \cite{oregan13} did random! Why was this changed?}The participants have not received any training tasks before the main evaluations. Therefore, it is possible that the accommodation period is responsible for lower-than-expected scores in this mode of evaluation.

@@

\begin{table*}
\centering
  \begin{tabular}{|l|r|r|r|r|}
  \hline
 \multirow{2}{*}{\textbf{Density}} & \multicolumn{4}{|c|}{Evaluation mode} \\\cline{2-5}
                                            & \textbf{MT \& Src} & \textbf{MT} & \textbf{Src} & \textbf{No hint} \\
\hline
10\%&0.598&0.459 &0.711 &0.517 \\
20\%&0.740&0.667 &0.473 &0.261 \\
30\%&0.534&0.581 &0.411 &0.412 \\
\hline
\end{tabular}
\caption {Tatar--Russian: Krippendorff Alpha measure} 
\label{table:alpha-tat-rus}
\end{table*}
We selected this measure because of its compatibility with more than two annotators per task and the missing data (not all the gaps were evaluated). To calculate Krippendorff's alpha we used an algorithm implementation by Thomas Grill,\footnote{http://grrrr.org/data/dev/krippendorff\_alpha/} dividing the answers in each gap into two categories: correct and incorrect. The previously obtained synonym lists were taken into account, i.e. if the two answers are different but both correct, they fall into one category. The measure was calculated separately for each hint and percentage combination.

As the MT and MT+source hints are introduced, the agreement between annotators increases (measures closer to 1): the annotators are more consistently correct or incorrect in each given sentence. The agreement measure for the same sentences without hints is closer to zero, which contributes to the reliability of our methodology. We note the outlier score in Tatar--Russian 10\% source tasks, which has the most contribution from the news texts (agreement of 0.917). This set of sentences contains many loan words, which have similar form in Tatar and Russian (e.g. president, minister, championship), and are understood by Russian speakers. The gaps with loan words have mostly been filled correctly, while there was some disagreement in other gaps.

% a little on categories
\subsection{Results for different domains}
For the Tatar--Russian language pair the participants were offered texts from three different domains (in equal proportions): casual conversations, legal texts and news. The results by domains are displayed in tables \ref{table:rus-casual}, \ref{table:rus-legal}, and \ref{table:rus-news}. The MT system used in the evaluation has been targeted to translate texts from all three of the domains. Taking into consideration the above discussion of 10\% MT+Source tasks, we observe similar results across the three categories. Note that, in casual and news texts, the source sentences paired with MT significantly improve participants' performance compared to MT-only task mode. This may be due to the fact that many words are borrowed from Russian into Tatar, and are in fact understood by Russian speakers.

@@

\begin{table*}
\centering
  \begin{tabular}{|l|r|r|r|r|}
  \hline
 \multirow{2}{*}{\textbf{Density}} & \multicolumn{4}{|c|}{Evaluation mode} \\\cline{2-5}
                                            & \textbf{MT \& Src} & \textbf{MT} & \textbf{Src} & \textbf{No hint} \\
\hline
10\%&62.19 \(\pm\) 32&58.04 \(\pm\) 28&39.90 \(\pm\) 39&48.79 \(\pm\) 40\\
20\%&64.81 \(\pm\) 30&70.18 \(\pm\) 27&31.37 \(\pm\) 28&30.91 \(\pm\) 30\\
30\%&48.13 \(\pm\) 26&39.55 \(\pm\) 24&25.94 \(\pm\) 20&18.07 \(\pm\) 18\\
\hline
\end{tabular}
\caption {Basque--Spanish: Average number of gaps successfully filled (\%), using a synonym list} \label{table:res-eus-spa}
\end{table*}

\begin{table*}
\centering
  \begin{tabular}{|l|r|r|r|r|}
    \hline
    \multirow{2}{*}{\textbf{Density}} & \multicolumn{4}{|c|}{Evaluation mode} \\\cline{2-5}
                                            & \textbf{MT \& Src} & \textbf{MT} & \textbf{Src} & \textbf{No hint} \\
\hline
10\%&63.76 \(\pm\) 41&57.02 \(\pm\) 42&53.52 \(\pm\) 43&45.96 \(\pm\) 41\\
20\%&59.77 \(\pm\) 33&64.93 \(\pm\) 31&45.60 \(\pm\) 31&38.56 \(\pm\) 32\\
30\%&55.83 \(\pm\) 26&58.77 \(\pm\) 28&40.00 \(\pm\) 28&35.45 \(\pm\) 30\\
    \hline
  \end{tabular}
  \caption{Tatar--Russian: Average number of gaps successfully filled (\%), using a synonym list}
  \label{table:res-tat-rus}
\end{table*}

@@

\centering
\begin{tabular}{|r |*{4}{c}|}
\hline
  &\multicolumn{4}{c|}{\textbf{Evaluation mode}}\\
\hline
\textbf{Gap percentage} & \textbf{MT \& Src} & \textbf{MT} & \textbf{Src} & \textbf{No hint} \\
\hline
10\%&63.64 \(\pm\) 45&63.64 \(\pm\) 43&62.12 \(\pm\) 46&53.03 \(\pm\) 44\\
20\%&73.33 \(\pm\) 32&63.33 \(\pm\) 36&41.11 \(\pm\) 28&37.78 \(\pm\) 31\\
30\%&70.28 \(\pm\) 31&59.58 \(\pm\) 24&39.44 \(\pm\) 27&37.71 \(\pm\) 33\\
\hline
\end{tabular}
\caption {Tatar--Russian, casual conversations: Average number of gaps successfully filled (\%), using a synonym list} \label{table:rus-casual} 
\end{table*}
